---
title: "Linear Regression Analysis- Utah Population Projection"
output:
  pdf_document: default
  html_document: default
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
#Load the Libraries
library(dplyr)
library(ggthemes)
library(ggplot2)
library(ggExtra)
library(readr)
library(tidyr)
library(ISLR) #mathematical models
library(caTools)
library(class)
library(e1071)
library(corrplot) 
library(knitr)
```


*Written by Savahnna L. Cunningham*

*Date: September 15, 2018*



```{r message=FALSE, warning=FALSE}
# Load the Data
UT_data<- read_csv("~/Desktop/MSDA Portfolio/2. DS Tools & Techniques/P2. R/ut_census.csv")

View(UT_data)
```

# Introduction

Conduct a linear regession analysis using the U.S. Census Bureau data for the state of Utah to predict the size of the population in 2020.

```{r message=FALSE, warning=FALSE}
# Split the dataset into the Training set and Test set
set.seed(123)
split = sample.split(UT_data$Pop_Est, SplitRatio = 2/3)
training_set = subset(UT_data, split == TRUE)
test_set = subset(UT_data, split == FALSE)

# Fit the Simple Linear Regression to the Training set
regressor = lm(formula = Pop_Est ~ Year,
               data = training_set)
```



```{r message=FALSE, warning=FALSE}
#Predict the Test set results
y_pred = predict.lm(regressor, newdata = test_set,interval = "prediction",level = 0.95)

#Visualize the result
smoothScatter(y_pred,pch = ".", cex = 5, 
                     col = "black",colramp = 
                     colorRampPalette(c("white", blues9)),
                     xlab = "Fit", 
                     ylab = "Model Prediction",
                     main="Predicted Future Values ")
```

The visualization represents the 95% prediction interval with data points<br> 
representing the models predicted values. As you can see, the model did very<br>
well predicting the population estimate, as all data points are within the<br>
prediction interval. 


```{r message=FALSE, warning=FALSE}
summary(regressor)
```


```{r message=FALSE, warning=FALSE}
#Confidence Interval for the Utah population estimate for the year 2020  
y_pred <- lm(regressor, data = test_set)
new_data <-data.frame(Year=2020)
utah_predict<- predict(y_pred,new_data,interval="predict",level=.95)
utah_predict
```

Using a 95% confidence interval, the results of the linear regression model indicate the estimated population for the state of Utah will be between 3,002,358 and 3,505,925 people. 

```{r message=FALSE, warning=FALSE}
#Utah state population estimate for the year 2020
utah_predict<- predict(y_pred,new_data)
utah_predict
```

The Linear Regression model result for the 2020 Utah state population is 3,254,141 people.


Population prediction for the state of Utah in 5 years time.


```{r message=FALSE, warning=FALSE}
#Confidence Interval for the Utah population estimate for the year 2023  
y_pred <- lm(regressor, data = test_set)
new_data <-data.frame(Year=2023)
utah_predict<- predict(y_pred,new_data,interval="predict",level=.95)
utah_predict
```

Using a 95% confidence interval, the results of the linear regression model indicate the estimated population for the state of Utah will be between 3,049,965 and 3,766,552 people. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
#Utah state population estimate for the year 2023
utah_predict<- predict(y_pred,new_data)
utah_predict
```

Predicted Utah state population for the year 2023 is 3,408,259 people.


